---
title: "Snake colour extraction — pipeline v3"
subtitle: "Chapter 3 (4) — background-referenced illuminant/exposure correction and k-means colour clustering"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 8
    fig-height: 4
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

## Overview

This document extracts, for every segmented snake photo, a **two-colour
summary** (the dorsal background colour and the darker zigzag colour) that is
comparable across photos taken under different illumination.

### What changed relative to v2

v2 only shifted the `L*` channel (exposure), leaving the colour cast (white
balance) in `a*`/`b*` untouched. v3 instead:

1. estimates the **scene illuminant** from the *background* pixels of the
   original GBIF photo (the mask is paired by filename, one directory up);
2. performs diagonal (**von Kries**) chromatic adaptation toward a neutral,
   D65-like reference;
3. normalises **exposure** to the scene's brightest non-saturated background
   pixels.

`a*`/`b*` values are therefore comparable across photos, while genuine snake
lightness variation is preserved (no forced `L*` target).

> **Relative, not absolute.** GBIF photos contain no in-image grey standard, so
> true reflectance is unrecoverable (Stevens et al. 2007; Troscianko & Stevens
> 2015). Metadata for every correction (WB gains, exposure gain, residual cast)
> is exported so downstream models (step 6) can absorb residual technical noise
> statistically.

The pipeline is deliberately **deterministic**: every stochastic step (background
subsampling, k-means subsampling and `nstart` restarts) is seeded from a hash of
the photo id, so re-running produces identical outputs.

------------------------------------------------------------------------

## 1. Libraries and parallel setup

```{r}
#| label: setup
#| message: false

library(imager)       # image IO and RGB <-> Lab conversion
library(dplyr)        # data wrangling
library(parallel)     # cluster creation
library(doParallel)   # cross-platform backend (PSOCK on Windows, fork on Linux)
library(foreach)      # explicit, so %dopar% is available even if not re-exported

# Leave headroom for the OS/UI: use (available cores - 2).
N_CORES <- max(1, parallel::detectCores() - 2)
```

## 2. File-system setup

Segmented snake crops (`*.png`, with alpha) live in `img_dir`; the matching
original GBIF photos (`*.jpg`) live one directory up and are used only to
estimate the background illuminant.

```{r}
#| label: paths

img_dir  <- "C:/Users/pdeschepper/OneDrive - Institute of Natural Sciences/Desktop/PERSONAL/DeepLearning/vipera_spatialcolors/Vipera_latastei_gbif_sourceimages/Extracted_snakes_pytorch"
orig_dir <- dirname(img_dir)   # background photos live one level up

images <- list.files(img_dir, pattern = "\\.png$", ignore.case = TRUE, full.names = TRUE)
stopifnot(length(images) > 0)

out_dir <- file.path(img_dir, "ColorExtraction_results_v3")
if (!dir.exists(out_dir)) dir.create(out_dir)

# Optional QC step: write out the corrected PNGs (snake + transparency
# preserved) so the correction can be inspected visually. Off by default
# because it doubles the disk footprint.
SAVE_CORRECTED_IMAGES <- FALSE
corr_dir <- NULL
if (SAVE_CORRECTED_IMAGES) {
  corr_dir <- file.path(img_dir, "Corrected")
  if (!dir.exists(corr_dir)) dir.create(corr_dir)
}
```

## 3. Correction parameters

These defaults are the analysis configuration; they are collected in one place
so the whole correction can be reproduced or sensitivity-tested.

```{r}
#| label: parameters

## ---- Illuminant estimation (Shades-of-Grey) ----------------------------- ##
SOG_P       <- 6      # p-norm exponent; Grey-World is p = 1, p = 6 is robust
                      # (Finlayson & Trezzi)
WP_WEIGHT   <- 0.30   # blend weight of the White-Patch estimate (rest -> Shades-of-Grey)
WP_QUANTILE <- 0.995  # "white" reference quantile within the background

## ---- Background quality filters ----------------------------------------- ##
BG_SAT_CUT    <- 0.98   # drop background pixels clipped in any channel
BG_BLACK_CUT  <- 0.02   # drop near-black pixels (sensor noise dominates)
BG_SAMPLE_MAX <- 20000  # background pixels subsampled per photo
BG_MIN_PIXELS <- 500    # minimum usable background pixels, else the image is skipped

## ---- Gain guard rails --------------------------------------------------- ##
# Avoid over-correcting atypical photos: clamp the gains and flag the photo
# instead of letting a bad estimate distort the colour data.
MIN_GAIN     <- 0.5
MAX_GAIN     <- 2.0
MIN_EXP_GAIN <- 0.2
MAX_EXP_GAIN <- 5.0

# Exposure normalisation: map the scene's brightest neutral background pixels
# (its strongest white reference) to this linear white level.
BG_WHITE_TARGET_LIN <- 0.92

## ---- Clustering (unchanged from v2) ------------------------------------- ##
SAMPLE_SIZE   <- 5000
N_CLUSTERS    <- 2
KMEANS_NSTART <- 10
```

## 4. Colour-conversion helpers

JPEG pixel values are gamma-encoded; von Kries diagonal adaptation is only
valid (approximately, for gamma-on images — Cardei & Funt 1999) in a linear
response space. These helpers convert between sRGB and linear light.

```{r}
#| label: colour-helpers

srgb_to_linear <- function(v) {
  ifelse(v <= 0.04045, v / 12.92, ((v + 0.055) / 1.055) ^ 2.4)
}

linear_to_srgb <- function(v) {
  ifelse(v <= 0.0031308, v * 12.92, 1.055 * v ^ (1 / 2.4) - 0.055)
}

# Shades-of-Grey channel statistic.
sog_stat <- function(x, p = SOG_P) {
  mean(x ^ p, na.rm = TRUE) ^ (1 / p)
}
```

## 5. Estimating the background correction

`estimate_background_correction()` works on the **original** photo. It:

1. takes the complement of the snake mask as background and subsamples it;
2. discards clipped and near-black pixels (no usable sensor response);
3. blends a **Shades-of-Grey** and a **White-Patch** illuminant estimate;
4. converts that estimate to per-channel von Kries gains, clamped to the guard
   rails;
5. derives a single exposure gain that maps the corrected background's white
   reference to `BG_WHITE_TARGET_LIN`.

It returns the gains plus QC metadata (saturation fraction, clamp flags, pixel
count) and the linearised background samples used later for the residual-cast
check.

```{r}
#| label: estimate-correction

estimate_background_correction <- function(orig, mask, seed) {

  set.seed(seed)

  orig_r <- as.vector(imager::channel(orig, 1))
  orig_g <- as.vector(imager::channel(orig, 2))
  orig_b <- as.vector(imager::channel(orig, 3))

  # Background = complement of the snake mask; subsample for speed.
  bg_idx <- which(!mask)
  if (length(bg_idx) < BG_MIN_PIXELS) {
    stop("Too little background for illuminant estimation.")
  }
  if (length(bg_idx) > BG_SAMPLE_MAX) {
    bg_idx <- bg_idx[sample.int(length(bg_idx), BG_SAMPLE_MAX)]
  }

  # Remove clipped and near-black pixels (no usable sensor response).
  enc_max <- pmax(orig_r[bg_idx], orig_g[bg_idx], orig_b[bg_idx])
  enc_min <- pmin(orig_r[bg_idx], orig_g[bg_idx], orig_b[bg_idx])
  usable  <- enc_max < BG_SAT_CUT & enc_min > BG_BLACK_CUT
  saturated_frac <- 1 - mean(usable)

  bg_idx <- bg_idx[usable]
  if (length(bg_idx) < BG_MIN_PIXELS) {
    stop("Too few usable background pixels after saturation/black filtering.")
  }

  r_lin <- srgb_to_linear(orig_r[bg_idx])
  g_lin <- srgb_to_linear(orig_g[bg_idx])
  b_lin <- srgb_to_linear(orig_b[bg_idx])

  # Illuminant estimate = weighted blend of Shades-of-Grey and White-Patch.
  sog_r <- sog_stat(r_lin); sog_g <- sog_stat(g_lin); sog_b <- sog_stat(b_lin)
  wp_r  <- stats::quantile(r_lin, probs = WP_QUANTILE, names = FALSE)
  wp_g  <- stats::quantile(g_lin, probs = WP_QUANTILE, names = FALSE)
  wp_b  <- stats::quantile(b_lin, probs = WP_QUANTILE, names = FALSE)

  est_r <- (1 - WP_WEIGHT) * sog_r + WP_WEIGHT * wp_r
  est_g <- (1 - WP_WEIGHT) * sog_g + WP_WEIGHT * wp_g
  est_b <- (1 - WP_WEIGHT) * sog_b + WP_WEIGHT * wp_b
  est   <- c(est_r, est_g, est_b)

  if (any(!is.finite(est)) || any(est <= 0)) {
    stop("Illuminant estimate is not finite/positive.")
  }

  # von Kries gains: scale channels so the estimated illuminant becomes neutral.
  neutral    <- mean(est)
  gains_raw  <- neutral / est
  gains      <- pmin(pmax(gains_raw, MIN_GAIN), MAX_GAIN)
  wb_clamped <- any(gains != gains_raw)

  # Exposure: white reference of the *corrected* background luminance.
  lum <- 0.2126 * (r_lin * gains[1]) + 0.7152 * (g_lin * gains[2]) + 0.0722 * (b_lin * gains[3])
  white_bg <- stats::quantile(lum, probs = WP_QUANTILE, names = FALSE)
  if (!is.finite(white_bg) || white_bg <= 0) {
    stop("Background white reference not usable.")
  }

  exp_gain_raw <- BG_WHITE_TARGET_LIN / white_bg
  exp_gain     <- pmin(pmax(exp_gain_raw, MIN_EXP_GAIN), MAX_EXP_GAIN)
  exp_clamped  <- !identical(exp_gain, exp_gain_raw)

  list(
    gains          = gains,
    exposure_gain  = exp_gain,
    saturated_frac = saturated_frac,
    wb_clamped     = wb_clamped,
    exp_clamped    = exp_clamped,
    n_bg_pixels    = length(bg_idx),
    bg_r = r_lin, bg_g = g_lin, bg_b = b_lin   # linearised, usable samples
  )
}
```

## 6. Applying the correction to the snake crop

The WB gains and the exposure gain are combined and applied in linear space;
the result is converted back to sRGB and to Lab for clustering.

```{r}
#| label: correct-crop

correct_crop <- function(img_crop, correction) {

  w  <- dim(img_crop)[1]
  h  <- dim(img_crop)[2]
  ch <- function(k) imager::channel(img_crop, k)

  g_r <- correction$gains[1] * correction$exposure_gain
  g_g <- correction$gains[2] * correction$exposure_gain
  g_b <- correction$gains[3] * correction$exposure_gain

  corr <- function(k, gain) {
    linear_to_srgb(pmin(pmax(srgb_to_linear(as.vector(k)) * gain, 0), 1))
  }

  new_r <- corr(ch(1), g_r)
  new_g <- corr(ch(2), g_g)
  new_b <- corr(ch(3), g_b)

  img_rgb <- imager::imappend(list(
    imager::as.cimg(new_r, dim = c(w, h, 1, 1)),
    imager::as.cimg(new_g, dim = c(w, h, 1, 1)),
    imager::as.cimg(new_b, dim = c(w, h, 1, 1))
  ), "c")

  list(
    rgb = img_rgb,
    lab = imager::RGBtoLab(img_rgb)
  )
}

# Residual cast of the corrected background samples in Lab. Values near 0 mean
# the neutral-background assumption holds; large values flag problem photos.
background_residual_cast <- function(correction) {
  r <- linear_to_srgb(pmin(pmax(correction$bg_r * correction$gains[1] * correction$exposure_gain, 0), 1))
  g <- linear_to_srgb(pmin(pmax(correction$bg_g * correction$gains[2] * correction$exposure_gain, 0), 1))
  b <- linear_to_srgb(pmin(pmax(correction$bg_b * correction$gains[3] * correction$exposure_gain, 0), 1))

  n <- length(r)
  bg_lab <- imager::RGBtoLab(imager::as.cimg(array(c(r, g, b), dim = c(n, 1, 1, 3))))
  a    <- stats::median(as.vector(imager::channel(bg_lab, 2)), na.rm = TRUE)
  bmed <- stats::median(as.vector(imager::channel(bg_lab, 3)), na.rm = TRUE)

  data.frame(residual_a = a, residual_b = bmed,
             residual_euclid = sqrt(a ^ 2 + bmed ^ 2))
}
```

## 7. Clustering into two colour zones

k-means on the corrected Lab pixels, masked to snake-only pixels and subsampled
for speed. Cluster ids are assigned by lightness `L*`, **not** by size:
cluster 1 = lighter (dorsal background), cluster 2 = darker (zigzag). This keeps
labels comparable across images regardless of which colour covers more area.

The per-photo `seed` is passed in so both the subsample and the k-means
restarts are reproducible.

```{r}
#| label: clustering

extract_lab_clusters <- function(std_result, n = N_CLUSTERS,
                                 sample.size = SAMPLE_SIZE,
                                 nstart = KMEANS_NSTART,
                                 seed = 1L) {

  lab_img <- std_result$lab
  mask    <- std_result$mask

  L <- as.vector(imager::channel(lab_img, 1))[mask]
  a <- as.vector(imager::channel(lab_img, 2))[mask]
  b <- as.vector(imager::channel(lab_img, 3))[mask]

  lab_matrix <- cbind(L, a, b)
  lab_matrix <- lab_matrix[stats::complete.cases(lab_matrix), , drop = FALSE]

  if (nrow(lab_matrix) < n) {
    stop("Not enough valid pixels to cluster.")
  }

  # Deterministic subsampling / initialisation (was set.seed(NULL), which made
  # the result irreproducible; now keyed to the photo id).
  set.seed(seed)
  if (nrow(lab_matrix) > sample.size) {
    lab_matrix <- lab_matrix[sample.int(nrow(lab_matrix), sample.size), ]
  }

  km <- stats::kmeans(lab_matrix, centers = n, nstart = nstart, iter.max = 50)

  # Order cluster ids from lightest (cluster 1, background) to darkest
  # (cluster 2, zigzag): lightest dorsal colour always has highest L*.
  ord <- order(km$centers[, 1], decreasing = TRUE)

  sizes   <- km$size / sum(km$size)
  centers <- km$centers[ord, , drop = FALSE]
  sizes   <- sizes[ord]

  rgb_vals <- grDevices::convertColor(centers, from = "Lab", to = "sRGB", clip = TRUE)
  rgb_vals[rgb_vals < 0] <- 0
  rgb_vals[rgb_vals > 1] <- 1
  hex <- grDevices::rgb(rgb_vals[, 1], rgb_vals[, 2], rgb_vals[, 3])

  data.frame(
    cluster    = seq_len(n),
    proportion = sizes,
    L = centers[, 1],
    a = centers[, 2],
    b = centers[, 3],
    hex = hex,
    stringsAsFactors = FALSE
  )
}
```

## 8. Per-image pipeline

`process_one_image()` ties the steps together for a single crop: load the crop
and its original, build the mask from alpha, estimate and apply the correction,
compute the residual cast, optionally save the corrected image, and cluster the
Lab pixels. A stable per-photo seed is derived from a hash of the photo id.

```{r}
#| label: process-one

process_one_image <- function(png_path) {

  photo_id <- tools::file_path_sans_ext(basename(png_path))
  jpg_path <- file.path(orig_dir, paste0(photo_id, ".jpg"))
  seed <- sum(utf8ToInt(photo_id)) %% 999983L   # deterministic subsampling per photo

  img_crop <- imager::load.image(png_path)
  has_alpha <- dim(img_crop)[4] == 4
  if (!has_alpha) stop("Cropped PNG has no alpha channel.")

  alpha <- imager::channel(img_crop, 4)
  mask  <- as.vector(alpha) > 0.5
  if (sum(mask) < 50) stop("Fewer than 50 snake pixels -- check the crop.")

  if (!file.exists(jpg_path)) stop("Original JPG not found for this photo.")
  img_orig <- imager::load.image(jpg_path)
  if (!identical(dim(img_crop)[1:2], dim(img_orig)[1:2])) {
    stop("Dimension mismatch between crop and original.")
  }

  correction <- estimate_background_correction(img_orig, mask, seed)
  corrected  <- correct_crop(img_crop, correction)
  residual   <- background_residual_cast(correction)

  if (SAVE_CORRECTED_IMAGES) {
    corrected_with_alpha <- imager::imappend(list(corrected$rgb, alpha), "c")
    imager::save.image(corrected_with_alpha, file.path(corr_dir, basename(png_path)))
  }

  std_result <- list(
    lab = corrected$lab,
    mask = mask,
    has_alpha = has_alpha
  )

  clust_df <- extract_lab_clusters(std_result, seed = seed)
  clust_df$photo_id <- photo_id
  clust_df$file     <- basename(png_path)

  meta <- data.frame(
    photo_id        = photo_id,
    wb_gain_r       = correction$gains[1],
    wb_gain_g       = correction$gains[2],
    wb_gain_b       = correction$gains[3],
    exposure_gain   = correction$exposure_gain,
    saturated_frac  = correction$saturated_frac,
    wb_clamped      = correction$wb_clamped,
    exp_clamped     = correction$exp_clamped,
    n_bg_pixels     = correction$n_bg_pixels,
    residual_a      = residual$residual_a,
    residual_b      = residual$residual_b,
    residual_euclid = residual$residual_euclid,
    stringsAsFactors = FALSE
  )

  list(clusters = clust_df, meta = meta)
}
```

## 9. Batch run over all images

The run is parallelised with `doParallel`. Because PSOCK workers start with an
empty workspace, every function and constant a worker needs is exported
explicitly via `.export`; `.packages` loads `imager` on each worker.

The cluster is stopped in a `finally` block so it cannot leak into the rest of
the session (important because Quarto keeps one R session for the whole
document).

```{r}
#| label: batch-run
#| message: false

cat(sprintf("Processing %d images on %d cores...\n", length(images), N_CORES))

# Per-job wrapper returning either a result or an error tag, so one bad image
# cannot abort the whole batch.
wrapper <- function(png_path) {
  tryCatch(
    process_one_image(png_path),
    error = function(e) list(error = sprintf("%s: %s", basename(png_path), conditionMessage(e)))
  )
}

# Everything the workers need (functions + constants), exported explicitly.
NEEDED_FOR_PARALLEL <- c(
  "wrapper", "process_one_image",
  "estimate_background_correction", "correct_crop",
  "background_residual_cast", "extract_lab_clusters",
  "srgb_to_linear", "linear_to_srgb", "sog_stat",
  "orig_dir", "SAVE_CORRECTED_IMAGES", "corr_dir",
  "N_CLUSTERS", "KMEANS_NSTART", "SAMPLE_SIZE",
  "BG_MIN_PIXELS", "BG_SAMPLE_MAX", "BG_SAT_CUT", "BG_BLACK_CUT",
  "WP_QUANTILE", "WP_WEIGHT", "SOG_P",
  "MIN_GAIN", "MAX_GAIN", "MIN_EXP_GAIN", "MAX_EXP_GAIN",
  "BG_WHITE_TARGET_LIN"
)

cl <- parallel::makeCluster(N_CORES, type = "PSOCK")
doParallel::registerDoParallel(cl)

t0 <- proc.time()

results <- tryCatch(
  foreach::foreach(
    png_path = images,
    .packages = "imager",
    .export = NEEDED_FOR_PARALLEL
  ) %dopar% {
    wrapper(png_path)
  },
  finally = {
    # Always release the workers and reset the foreach backend.
    parallel::stopCluster(cl)
    foreach::registerDoSEQ()
  }
)

cat(sprintf("Elapsed: %.1f min\n", (proc.time() - t0)["elapsed"] / 60))
```

```{r}
#| label: batch-outputs
#| message: false

is_error <- vapply(results, function(r) !is.null(r$error), logical(1))
good     <- results[!is_error]
failed   <- vapply(results[is_error], function(r) r$error, character(1))

clusters_df <- dplyr::bind_rows(lapply(good, function(r) r$clusters)) %>%
  dplyr::select(photo_id, file, cluster, proportion, L, a, b, hex)

meta_df <- dplyr::bind_rows(lapply(good, function(r) r$meta))

# Per-cluster CSV (response data) plus per-photo correction CSV (covariates).
write.csv(clusters_df, file.path(out_dir, "snake_color_clusters_v3.csv"), row.names = FALSE)
write.csv(meta_df,     file.path(out_dir, "correction_metadata.csv"),    row.names = FALSE)

if (length(failed) > 0) {
  cat("\nThe following images failed and were skipped:\n")
  print(failed)
  writeLines(failed, file.path(out_dir, "failed_images_v3.txt"))
}

cat(sprintf(
  "\nDone. Processed %d/%d images successfully. Results written to:\n%s\n",
  length(good), length(images), out_dir
))
```

## 10. QC summary and visual inspection

The summaries below check that the correction did not run away: gains should sit
inside the guard rails, and the corrected background residual cast (`a*`, `b*`)
should centre near 0.

```{r}
#| label: qc-summary

cat("\nCorrection gain distribution (all photos):\n")
print(summary(meta_df[, c("wb_gain_r", "wb_gain_g", "wb_gain_b", "exposure_gain")]))

cat("\nBackground residual cast (a*, b*): should centre near 0\n")
print(summary(meta_df[, c("residual_a", "residual_b")]))

cat(sprintf("\nImages with clamped WB gains or exposure: %.1f%%\n",
            100 * mean(meta_df$wb_clamped | meta_df$exp_clamped)))
```

### Side-by-side inspection

`compare_correction()` renders the original crop next to the corrected crop for
one photo. This chunk is **not executed on render** — set `eval: true` (or run
it interactively) to inspect specific photos.

```{r}
#| label: compare-correction
#| eval: false

compare_correction <- function(png_path) {
  original <- imager::load.image(png_path)
  alpha    <- imager::channel(original, 4)
  jpg_path <- file.path(orig_dir, paste0(tools::file_path_sans_ext(basename(png_path)), ".jpg"))

  seed <- sum(utf8ToInt(tools::file_path_sans_ext(basename(png_path)))) %% 999983L
  correction <- estimate_background_correction(imager::load.image(jpg_path),
                                               as.vector(alpha) > 0.5, seed)
  corrected_img_rgb <- correct_crop(original, correction)$rgb
  corrected <- imager::imappend(list(corrected_img_rgb, alpha), "c")

  old_par <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
  plot(original,  axes = FALSE, main = "Original")
  plot(corrected, axes = FALSE, main = "Corrected")
  par(old_par)
}

# Example inspections:
# lapply(images[500:510], compare_correction)
compare_correction(images[14])
```

```{r}
#| label: files-written

list.files(out_dir)
```

```{r}
#| label: session-info
#| echo: false

sessionInfo()
```
