---
title: "Environment × colour association in *Vipera aspis*"
subtitle: "Chapter 3 (6b) — confounder-aware spatial GAMs and permutation tests"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 10
    fig-height: 8
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
  cache: false
---

## Overview

This notebook links colour polymorphism in *Vipera sp* (background and
zigzag colour, quantified in pipeline `4 … v3`) to the local environment at the
**observation level**, while controlling for the technical noise inherent in
opportunistic GBIF / iNaturalist photographs.

It answers two questions:

1. **Which colour trait responds to which environmental driver, how strongly,
   and in which direction?**
   → confounder-aware **GAMs** (`mgcv`) fitted to four crystallised colour
   traits (per-zone lightness, zigzag–background contrast, warmth), with a
   spatial Gaussian-process smooth, subspecies, season and image-technical
   covariates as noise controls.
2. **Is colour variation non-randomly arranged in space — including localised
   "checkerboards" that a Mantel gradient test would miss?**
   → a **permutation test** of Moran's *I* plus a neighbour trait-contrast
   statistic on a k-nearest-neighbour graph.

> **Photographer / camera caveat.** iNaturalist no longer exposes photographer
> identity, so no random-effect grouping is possible there. Remaining per-photo
> technical noise is handled by (i) the v3 correction covariates (WB gains,
> exposure, residual cast), (ii) inverse-variance quality weighting built from
> within-observation photo dispersion, and (iii) robust interpretation: effect
> sizes are attenuated by measurement error, so the environmental signal is a
> **conservative lower bound**.

> **Subspecies caveat.** Subspecies are assigned from an editable geographic
> lookup table (`SUBSPECIES_LUT`). **Verify the boxes against the species range
> and current taxonomy before trusting the subspecies term.**

------------------------------------------------------------------------

## 0. Setup and configuration

All package dependencies and the machine-specific paths are collected in one
place. Paths are the only thing you should need to edit to run this on another
machine.

```{r}
#| label: setup
#| message: false

library(dplyr)      # data wrangling
library(tidyr)      # pivot / unnest
library(stringr)    # text parsing of the GBIF photo-id list
library(sf)         # spatial points and distances
library(terra)      # raster extraction fallback
library(mgcv)       # GAMs, REML/ML, shrinkage smoothers
library(farver)     # CIEDE2000 perceptual colour difference
library(ggplot2)    # all figures
library(lubridate)  # observation_date -> day-of-year

# Optional: usdm provides a vetted VIF-step procedure. A built-in fallback
# with the same pruning rule is used if usdm is not installed.
if (requireNamespace("usdm", quietly = TRUE)) library(usdm)
```

```{r}
#| label: config-paths
#| message: false
## ---- Paths (EDIT to your machine) --------------------------------------- ##
base_dir    <- "C:/Users/pdeschepper/OneDrive - Institute of Natural Sciences/Desktop/PERSONAL/DeepLearning/vipera_spatialcolors"
species_dir <- file.path(base_dir, "Vipera_aspis_gbif_sourceimages")
snake_dir   <- file.path(species_dir, "Extracted_snakes_pytorch")
res_dir     <- file.path(snake_dir, "ColorExtraction_results_v3")

clusters_csv <- file.path(res_dir, "snake_color_clusters_v3.csv")
corr_csv     <- file.path(res_dir, "correction_metadata.csv")
coverage_csv <- file.path(snake_dir, "class_coverage.csv")
gbif_csv     <- file.path(species_dir, "Vipera_aspis_gbif_metadata.csv")

ENVROOT <- "C:/Users/pdeschepper/OneDrive - Institute of Natural Sciences/Desktop/PERSONAL/Land_dataproducts/"

out_dir <- file.path(base_dir, "MLM_results_Vipera_aspis")
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
```

```{r}
#| label: config-params
#| message: false

## ---- Analysis parameters ------------------------------------------------- ##
COVERAGE_THRESHOLD <- 10   # % class coverage below which a photo is dropped (as in step 5)
ENV_SMOOTH_K       <- 4    # per-environment smooth basis size (small; shrinkage smoother)
SPATIAL_K          <- 80   # GP smooth basis size for s(longitude, latitude)
PERM               <- 999  # permutations for the spatial randomness test

# Fix the RNG once, at the top, so permutations are reproducible; the
# permutation test also calls set.seed() internally for extra safety.
set.seed(42)
```

------------------------------------------------------------------------

## 1. Colour data at the photo level

Three inputs are joined on `photo_id`:

* `snake_color_clusters_v3.csv` — the two colour clusters per photo (cluster 1
  = lighter background, cluster 2 = darker zigzag) with their CIELab values and
  pixel proportions;
* `correction_metadata.csv` — the technical corrections estimated in pipeline 4
  v3 (white-balance gains, exposure, residual cast, saturation);
* `class_coverage.csv` — the segmentation class coverage, used as a quality
  filter.

The GBIF metadata stores the linked photo ids as a Python-style string
(`"['a.jpg', 'b.jpg']"`), which is parsed into one row per photo.

```{r}
#| label: load-photo-data
#| message: false

clusters <- read.csv(clusters_csv, stringsAsFactors = FALSE)
corr_md  <- read.csv(corr_csv,     stringsAsFactors = FALSE)
gbif     <- read.csv(gbif_csv,     stringsAsFactors = FALSE)

stopifnot(c("photo_id", "cluster", "L", "a", "b") %in% names(clusters))
stopifnot(c("photo_id", "wb_gain_r") %in% names(corr_md))

# One row per photo: pivot the two colour clusters (1 = background,
# 2 = zigzag) into wide L/a/b + proportion columns.
clusters_wide <- clusters %>%
  distinct(photo_id, cluster, .keep_all = TRUE) %>%
  dplyr::select(photo_id, cluster, proportion, L, a, b) %>%
  tidyr::pivot_wider(
    id_cols    = photo_id,
    names_from = cluster,
    values_from = c(proportion, L, a, b),
    names_glue = "{.value}_{cluster}"
  ) %>%
  dplyr::filter(
    !is.na(L_1), !is.na(L_2), !is.na(a_1),
    !is.na(b_1), !is.na(a_2), !is.na(b_2)
  )

# Class coverage (photo_ID column, extension stripped so ids match)
coverage <- read.csv(coverage_csv, stringsAsFactors = FALSE) %>%
  mutate(photo_id = stringr::str_remove(photo_ID, "\\.[A-Za-z]+$")) %>%
  dplyr::select(photo_id, class_coverage)

# GBIF observation -> photo link table (parse the Python-list string)
gbif_long <- gbif %>%
  mutate(photo_id_list = stringr::str_extract_all(photo_ids, "'([^']+)'")) %>%
  tidyr::unnest(photo_id_list) %>%
  mutate(photo_id = stringr::str_remove_all(photo_id_list, "'")) %>%
  dplyr::select(gbif_id, latitude, longitude, observation_date, photo_id) %>%
  distinct()

photo_level <- clusters_wide %>%
  inner_join(coverage,  by = "photo_id") %>%
  inner_join(corr_md,   by = "photo_id") %>%
  inner_join(gbif_long, by = "photo_id") %>%
  dplyr::filter(class_coverage > COVERAGE_THRESHOLD)

cat(nrow(photo_level), "photos passed the coverage filter.\n")
```

------------------------------------------------------------------------

## 2. Aggregate to the observation level (with quality weights)

Colour is averaged in CIELab **per zone** across all photos of an observation.
Quality is encoded in a weight `w` built from two ingredients:

* **informative mass** — `n_photos × mean class coverage`, i.e. how much
  reliable segmentation evidence the observation carries;
* **technical noise** — the within-observation dispersion (SD) of the six Lab
  values across photos. Observations whose photos disagree are down-weighted.

Because photographer identity is unavailable, these weights plus the v3
correction covariates carry the technical-noise information. Weights are
rescaled to mean 1 so they do not change the scale of the model coefficients.

```{r}
#| label: obs-level
#| message: false

obs_level <- photo_level %>%
  group_by(gbif_id, latitude, longitude, observation_date) %>%
  summarise(
    n_photos   = n(),
    prop_1     = mean(proportion_1, na.rm = TRUE),
    prop_2     = mean(proportion_2, na.rm = TRUE),

    # Zone means, weighted by the pixel proportion of each colour zone.
    L_1 = weighted.mean(L_1, w = proportion_1, na.rm = TRUE),
    a_1 = weighted.mean(a_1, w = proportion_1, na.rm = TRUE),
    b_1 = weighted.mean(b_1, w = proportion_1, na.rm = TRUE),
    L_2 = weighted.mean(L_2, w = proportion_2, na.rm = TRUE),
    a_2 = weighted.mean(a_2, w = proportion_2, na.rm = TRUE),
    b_2 = weighted.mean(b_2, w = proportion_2, na.rm = TRUE),

    mean_class_coverage  = mean(class_coverage,    na.rm = TRUE),
    mean_wb_gain_r       = mean(wb_gain_r,         na.rm = TRUE),
    mean_wb_gain_g       = mean(wb_gain_g,         na.rm = TRUE),
    mean_wb_gain_b       = mean(wb_gain_b,         na.rm = TRUE),
    mean_exposure_gain   = mean(exposure_gain,     na.rm = TRUE),
    mean_residual_euclid = mean(residual_euclid,   na.rm = TRUE),
    mean_saturated_frac  = mean(saturated_frac,    na.rm = TRUE),

    # Technical-noise proxy: mean SD of the six Lab values across photos.
    dispersion = mean(c(
      stats::sd(L_1, na.rm = TRUE), stats::sd(a_1, na.rm = TRUE),
      stats::sd(b_1, na.rm = TRUE), stats::sd(L_2, na.rm = TRUE),
      stats::sd(a_2, na.rm = TRUE), stats::sd(b_2, na.rm = TRUE)
    ), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    doy = lubridate::yday(as.Date(substr(observation_date, 1, 10))),

    # dispersion is undefined for single-photo observations (SD of one value
    # is NA). Impute those from the median of multi-photo observations; if none
    # exist, fall back to a floor of 1 CIELab unit so dispersion can never be NA.
    dispersion = ifelse(!is.finite(dispersion),
                        stats::median(dispersion, na.rm = TRUE), dispersion),
    dispersion = ifelse(!is.finite(dispersion), 1, dispersion),

    inform = n_photos * pmin(mean_class_coverage / 100, 1),
    inform = ifelse(!is.finite(inform) | inform < 1, 1, inform),

    # inverse-variance-style weight: informative mass penalised by tech noise
    w = inform / (dispersion^2 + 0.01),
    w = ifelse(!is.finite(w) | w <= 0, 1, w)
  )

# Normalise weights to mean 1 so model scales are unaffected.
obs_level$w <- obs_level$w / mean(obs_level$w, na.rm = TRUE)

cat(sprintf("Observation-level weight diagnostics: %d obs, %d with >1 photo,\n",
            nrow(obs_level), sum(obs_level$n_photos > 1)))
cat(sprintf("  weight range %.3f - %.3f (mean 1), %d NA\n",
            min(obs_level$w, na.rm = TRUE), max(obs_level$w, na.rm = TRUE),
            sum(is.na(obs_level$w))))

if (any(is.na(obs_level$w))) {
  obs_level$w[is.na(obs_level$w)] <- 1
  cat("  (imputed residual NA weights to 1)\n")
}
```

------------------------------------------------------------------------

## 3. Environmental covariates, VIF pruning and subspecies

### 3a. Load point-level environment at the observations

The code extracts land cover, lithology, the 19 bioclim layers, elevation and
the 13-zone EnZ environmental stratification (Metzger et al. 2018) from the raw
rasters under `ENVROOT`. Any step-6 trait columns or duplicate
coordinate/date columns in the cache are dropped before joining, and the join
is keyed on `gbif_id` (coerced to character on both sides so a type mismatch
cannot silently produce zero matches).

```{r}
#| label: env-load
#| message: false

load_env_at_points <- function(occ_sf, envroot) {

  pts <- terra::vect(occ_sf)

  # Paths
  land_cover_path <- file.path(
    envroot,
    "Corine_landcover_vector_2018/103580/Results/",
    "u2018_clc2018_v2020_20u1_raster100m/",
    "u2018_clc2018_v2020_20u1_raster100m/DATA/",
    "U2018_CLC2018_V2020_20u1.tif"
  )

  lithology_path <- file.path(
    envroot, "EGDI_europe_lithology/GeologicUnitView.gpkg"
  )

  bioclim_path <- file.path(
    envroot, "bioclim/climate/wc2.1_2.5m"
  )

  elevation_path <- file.path(
    envroot,
    "elevation/elevation/wc2.1_2.5m/wc2.1_2.5m_elev.tif"
  )

  enz_path <- file.path(
    envroot,
    "Metzger_Environmental_stratification_of_Europe(EnZ)/ens_v8.shp"
  )

  # Raster extraction helper
  extract_raster <- function(x, pts, name = NULL) {

    r <- if (inherits(x, "SpatRaster")) x else terra::rast(x)

    if (!terra::hasValues(r))
      stop("Raster has no values: ", paste(names(r), collapse = ", "))

    p <- terra::project(pts, terra::crs(r))

    out <- terra::extract(r, p, ID = FALSE)
    out <- as.data.frame(out)

    if (!is.null(name) && ncol(out) == 1)
      names(out) <- name

    out
  }

  # Land cover
  land_cover <- extract_raster(
    land_cover_path, pts, "land_cover"
  )

  # Bioclim
  bio_files <- list.files(
    bioclim_path,
    pattern = "\\.tif$",
    full.names = TRUE
  )

  if (!length(bio_files))
    stop("No bioclim .tif files found in: ", bioclim_path)

  bioclim <- extract_raster(
    terra::rast(bio_files), pts
  )

  # Elevation
  elevation <- extract_raster(
    elevation_path, pts, "elevation"
  )

  # EnZ / EnS
  ens <- terra::vect(enz_path)
  p <- terra::project(pts, terra::crs(ens))

  enz <- terra::extract(
    ens[, c("EnS", "EnS_name", "EnZ", "EnZ_name")],
    p
  )
  enz <- enz[, c("EnS_name", "EnZ_name")] # remove ID column

  # Lithology
  lv <- terra::vect(lithology_path)
  p <- terra::project(pts, terra::crs(lv))

  lith <- terra::extract(lv, p)

  lithology <- data.frame(
    lithology = lith$representativelithology_title[
      match(seq_len(nrow(pts)), lith[[1]])
    ]
  )

  # Combine
  cbind(
    land_cover,
    lithology,
    bioclim,
    elevation,
    enz
  )
}

occ_sf_full <- obs_level %>%
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

env_raw <- load_env_at_points(occ_sf = occ_sf_full, envroot = ENVROOT)
dat <- bind_cols(obs_level, env_raw)


# Safety net: if any .x/.y duplicates slipped through, keep the obs_level copy
# and drop the .y one. dat$latitude / dat$longitude must exist as plain columns
# for assign_subspecies(), the maps and the models.
dupe_y <- intersect(c("latitude.y", "longitude.y", "observation_date.y"), names(dat))
if (length(dupe_y)) {
  cat("(clean-up) dropping duplicated join columns:", paste(dupe_y, collapse = ", "), "\n")
  dat <- dat %>% dplyr::select(-all_of(dupe_y))
}
```

### 3b. VIF pruning of the bioclimatic layers

The 19 WorldClim bioclim variables are highly collinear. We drop variables
iteratively until all remaining VIFs are below 5, using `usdm::vifstep()` when
available and an equivalent built-in step otherwise. The survivors are renamed
to safe syntactic names (`bio_1`, `bio_2`, …) for formula use.

```{r}
#| label: env-vif
#| message: false

bioclim_names <- grep("bio", names(dat), value = TRUE)

vif_select <- function(x, th = 5) {

  if (requireNamespace("usdm", quietly = TRUE)) {
    v <- usdm::vifstep(as.data.frame(x), th = th)
    return(as.character(v@results$Variables))
  }

  # Built-in fallback (same pruning rule: drop the highest VIF, repeat).
  keep <- names(x)
  repeat {
    vif <- 1 / (1 - sapply(keep, function(k) {
      z <- x[, setdiff(keep, k), drop = FALSE]
      if (ncol(z) == 0) return(0)
      summary(stats::lm(x[[k]] ~ ., data = as.data.frame(z)))$r.squared
    }))
    if (max(vif) <= th || length(keep) <= 2) break
    keep <- keep[-which.max(vif)]
  }
  keep
}

complete_rows <- stats::complete.cases(dat[, bioclim_names, drop = FALSE])
bio_sel <- vif_select(dat[complete_rows, bioclim_names, drop = FALSE], th = 5)

cat("Bioclim retained after VIF pruning (", length(bio_sel), "of", length(bioclim_names), "):\n")
print(bio_sel)

# Rename to safe syntactic names for formulas.
bio_sel    <- bio_sel[bio_sel %in% names(dat)]
bio_rename <- stats::setNames(paste0("bio_", seq_along(bio_sel)), bio_sel)
idx        <- match(names(bio_rename), names(dat))
names(dat)[idx] <- as.character(bio_rename)

continuous_vars <- c("elevation", paste0("bio_", seq_along(bio_rename)))
continuous_vars <- continuous_vars[continuous_vars %in% names(dat)]
if (!"elevation" %in% continuous_vars) cat("(warn) 'elevation' not found in env data.\n")
cat("Continuous env variables for analysis:", paste(continuous_vars, collapse = ", "), "\n")
```

### 3c. Subspecies from an editable geographic lookup

**Placeholder boxes — edit and verify** against the species range and current
taxonomy. You may instead supply your own table as a CSV with columns
`name, lon_min, lon_max, lat_min, lat_max, priority`. Points outside all boxes
stay `NA` and are dropped before modelling.

```{r}
#| label: subspecies
#| message: false

SUBSPECIES_LUT <- data.frame(
  name     = c("zinnikeri", "hugyi",   "francisciredi", "aspis"),
  lon_min  = c(-3.0,        13.5,       6.2,            -6.0),
  lon_max  = c(3.5,         18.5,       15.0,            8.0),
  lat_min  = c(40.5,        37.0,       43.2,           42.0),
  lat_max  = c(44.8,        43.5,       47.5,           51.0),
  priority = c(1, 2, 3, 4),
  stringsAsFactors = FALSE
)

# Optional CSV override (same columns as SUBSPECIES_LUT).
SUBSPECIES_LUT_CSV <- NULL

resolve_lut <- function() {
  lut <- SUBSPECIES_LUT
  if (!is.null(SUBSPECIES_LUT_CSV) && file.exists(SUBSPECIES_LUT_CSV)) {
    lut <- read.csv(SUBSPECIES_LUT_CSV, stringsAsFactors = FALSE)
    lut <- lut[order(lut$priority), ]
  }
  lut
}

# Assign each point to the first (highest-priority) box that contains it.
assign_subspecies <- function(dat, lut = resolve_lut()) {
  s <- rep(NA_character_, nrow(dat))
  for (i in seq_len(nrow(lut))) {
    hit <- is.na(s) &
      dat$latitude  >= lut$lat_min[i] & dat$latitude  <= lut$lat_max[i] &
      dat$longitude >= lut$lon_min[i] & dat$longitude <= lut$lon_max[i]
    s[hit] <- lut$name[i]
  }
  s
}

subspecies_lut    <- resolve_lut()
dat$subspecies    <- assign_subspecies(dat)

# Per-observation assignment, written out for external checking (e.g. QGIS).
if ("gbif_id" %in% names(dat)) {
  write.csv(dat[, c("gbif_id", "longitude", "latitude", "subspecies")],
            file.path(out_dir, "subspecies_points.csv"), row.names = FALSE)
}

cat("\nSubspecies assignment (VERIFY these counts!):\n")
print(table(dat$subspecies, useNA = "ifany"))
write.csv(data.frame(table(dat$subspecies, useNA = "ifany")),
          file.path(out_dir, "subspecies_assignment.csv"))
```

```{r}
#| label: subspecies-map
#| fig-cap: "Assumed subspecies boxes (dashed) over the actual observations."

# Range boxes vs. actual observations, so a mis-specified box is visible.
plot_subspecies_map <- function(dat, lut, xlim = c(-8, 22), ylim = c(34, 52)) {
  cols <- c(zinnikeri = "#1b9e77", aspis = "#7570b3",
            francisciredi = "#d95f02", hugyi = "#e7298a")

  p <- ggplot2::ggplot()
  if (requireNamespace("maps", quietly = TRUE)) {
    p <- p + ggplot2::borders("world", xlim = xlim, ylim = ylim,
                              colour = "grey40", fill = "grey90")
  }

  pts <- dat[!is.na(dat$subspecies), c("longitude", "latitude", "subspecies"), drop = FALSE]

  p +
    ggplot2::geom_point(data = pts,
                        ggplot2::aes(longitude, latitude, fill = subspecies),
                        shape = 21, colour = "grey15", size = 1.5, alpha = 0.85) +
    ggplot2::geom_rect(data = lut,
                       ggplot2::aes(xmin = lon_min, xmax = lon_max,
                                    ymin = lat_min, ymax = lat_max, colour = name),
                       fill = NA, linetype = "dashed") +
    ggplot2::scale_fill_manual(values = cols, name = "Assumed subspecies") +
    ggplot2::scale_colour_manual(values = cols, name = "Assumed subspecies") +
    ggplot2::coord_quickmap(xlim = xlim, ylim = ylim) +
    ggplot2::labs(x = "longitude", y = "latitude") +
    ggplot2::theme_minimal()
}

plot_subspecies_map(dat, subspecies_lut)
```

### 3d. Final analysis dataset and crystallised colour traits

Four response traits are derived per observation:

| Trait | Definition | Meaning |
|:--|:--|:--|
| `L_1` | lightness of cluster 1 | background lightness |
| `L_2` | lightness of cluster 2 | zigzag lightness |
| `contrast` | CIEDE2000 distance between the two zone colours | pattern contrast |
| `warmth` | `b*_1 − a*_1` | background yellowness (Cartesian, model-safe) |

Rare categorical levels (`< 5` observations) of land cover, lithology and EnZ
are pooled into `"other"` so the model matrix cannot explode with sparse
dummies.

```{r}
#| label: final-dataset
#| message: false

lab_cols  <- c("L_1", "a_1", "b_1", "L_2", "a_2", "b_2")
tech_vars <- c("mean_wb_gain_r", "mean_wb_gain_g", "mean_wb_gain_b",
               "mean_exposure_gain", "mean_residual_euclid", "mean_saturated_frac")

keep_cols <- unique(c("gbif_id", "longitude", "latitude", "doy", "w",
                      lab_cols, "prop_1", "prop_2", continuous_vars,
                      "land_cover", "lithology", "EnZ_name", "EnS_name", "subspecies", tech_vars))

dat <- dat[, keep_cols, drop = FALSE] %>%
  mutate(across(any_of(c(continuous_vars, lab_cols, tech_vars)), ~ as.numeric(.x))) %>%
  filter(!is.na(latitude), !is.na(longitude))

# Pool rare levels of the categorical covariates.
pool_rare_levels <- function(x, min_n = 5, what = "") {
  xc <- as.character(x)
  k  <- table(xc)
  small <- names(k)[k < min_n & !is.na(names(k)) & names(k) != "other"]
  if (length(small)) {
    cat(sprintf("  (pool) %s: %d class(es) with < %d obs -> 'other'\n",
                what, length(small), min_n))
    xc[xc %in% small] <- "other"
  }
  factor(xc)
}

for (v in c("land_cover", "lithology", "EnZ")) {
  if (v %in% names(dat)) dat[[v]] <- pool_rare_levels(dat[[v]], what = v)
}

# --- Derived traits: the crystallised response set -------------------------
# L_1 / L_2  : lightness of the lighter (background) and darker (zigzag) zone.
# contrast    : perceptual CIEDE2000 colour difference between the two zones.
# warmth      : background yellowness, b*_1 - a*_1 -- Cartesian, so it can be
#               used directly in models (unlike the circular hue angle).
tmp_ct <- farver::compare_colour(
  as.matrix(dat[, c("L_1", "a_1", "b_1"), drop = FALSE]),
  as.matrix(dat[, c("L_2", "a_2", "b_2"), drop = FALSE]),
  from_space = "lab", to_space = "lab", method = "cie2000"
)
dat$contrast <- diag(tmp_ct)
dat$warmth   <- dat$b_1 - dat$a_1
trait_cols   <- c("L_1", "L_2", "contrast", "warmth")

cat("Crystallised colour traits:", paste(trait_cols, collapse = ", "), "\n")

# --- Complete cases and dropping of unassigned points ----------------------
n_na <- colSums(is.na(dat))
if (any(n_na > 0)) {
  cat("\nNA counts by column (before complete-cases):\n")
  print(n_na[n_na > 0])
}

cc <- stats::complete.cases(dat)
cat(sprintf("\nObservations retained after complete-cases: %d / %d\n", sum(cc), nrow(dat)))
dat_full <- dat
dat      <- dat[cc, ]

if (!nrow(dat)) {
  cat("\n--- Column NA / constantness diagnostics ---\n")
  na_pct <- sort(colMeans(is.na(dat_full)) * 100, decreasing = TRUE)
  print(round(na_pct, 1))
  stop("All observations were dropped by complete.cases -- inspect the NA diagnostic above.")
}

n_unk <- sum(is.na(dat$subspecies))
if (n_unk > 0) {
  cat(sprintf("  -> dropping %d observation(s) with unassigned subspecies.\n", n_unk))
  dat <- dat[!is.na(dat$subspecies), ]
}
```

------------------------------------------------------------------------

## 4. Confounder-aware spatial GAMs

For each colour trait we fit a **full** and a **reduced** model with `mgcv`:

* **Full model** — environmental smooths (`bs = "ts"`, shrinkage) **+** a spatial
  Gaussian-process smooth `s(longitude, latitude, bs = "gp")` **+** a cyclic
  day-of-year smooth `s(doy, bs = "cc")` **+** subspecies / land cover /
  lithology / EnZ **+** the technical correction covariates.
* **Reduced model** — everything except the environmental smooths.

This isolates the *environmental* contribution above and beyond geography,
season, subspecies and image technology. `select = TRUE` adds an extra
shrinkage penalty so genuinely uninformative smooths collapse to ~0 edf. The
environmental block is tested with an approximate Chi-square LRT using **ML**
fits (REML is invalid for comparing models with different fixed effects); R²
and deviance explained are reported from the **REML** fits.

```{r}
#| label: gam-fit
#| message: false

fit_trait_models <- function(
    d,
    traits   = NULL,
    env_vars = NULL,
    cat_vars = c("land_cover", "lithology", "EnZ", "subspecies"),
    tech_vars = NULL,
    k_env = ENV_SMOOTH_K,
    k_sp  = SPATIAL_K
) {

  # Resolve globals from the calling frame rather than as recursive defaults.
  if (is.null(traits))    traits    <- get("trait_cols",      envir = parent.frame())
  if (is.null(env_vars))  env_vars  <- get("continuous_vars", envir = parent.frame())
  if (is.null(tech_vars)) tech_vars <- get("tech_vars",      envir = parent.frame())

  if (nrow(d) < 30) {
    stop(sprintf("Spatial GAM needs >= 30 rows; got %d.", nrow(d)))
  }

  # Keep only usable predictors (present, non-constant); otherwise mgcv raises
  # errors such as "contrasts can be applied only to factors with 2+ levels".
  env_vars  <- env_vars[env_vars %in% names(d)]
  cat_vars  <- cat_vars[cat_vars %in% names(d)]
  tech_vars <- tech_vars[tech_vars %in% names(d)]

  cat_vars <- cat_vars[vapply(cat_vars, function(v)
    length(unique(d[[v]][!is.na(d[[v]])])) > 1, logical(1))]

  env_vars <- env_vars[vapply(env_vars, function(v) {
    x <- d[[v]]
    length(unique(x[!is.na(x)])) > 1
  }, logical(1))]

  if (!length(env_vars)) stop("GAM: no non-constant environmental variables available.")

  # Character predictors -> factors.
  for (v in c(cat_vars, tech_vars)) {
    if (v %in% names(d) && is.character(d[[v]])) d[[v]] <- factor(d[[v]])
  }

  if (!"w" %in% names(d)) d$w <- 1

  cat("  (GAM) environmental variables:", paste(env_vars, collapse = ", "), "\n")

  # Shared (non-environmental) terms: space, season, categoricals, technology.
  base_terms <- c(
    sprintf("s(longitude, latitude, bs = 'gp', k = %d)", k_sp),
    "s(doy, k = 4, bs = 'cc')",
    cat_vars,
    tech_vars
  )

  env_terms <- vapply(env_vars, function(v)
    sprintf("s(%s, k = %d, bs = 'ts')", v, k_env), character(1))

  full_form <- function(trait)
    as.formula(paste(trait, "~", paste(c(env_terms, base_terms), collapse = " + ")))

  red_form <- function(trait)
    as.formula(paste(trait, "~", paste(base_terms, collapse = " + ")))

  rows <- lapply(traits, function(tr) {

    # Complete cases across every variable this model needs.
    vars_needed <- unique(c(tr, env_vars, cat_vars, tech_vars,
                            "longitude", "latitude", "doy", "w"))
    dd <- d[complete.cases(d[, vars_needed, drop = FALSE]), , drop = FALSE]

    if (nrow(dd) < 30) {
      stop(sprintf("Trait '%s' has fewer than 30 complete observations (%d).", tr, nrow(dd)))
    }

    # Final (REML) models.
    f_full <- mgcv::gam(full_form(tr), data = dd, method = "REML",
                        weights = w, select = TRUE, na.action = na.exclude)
    f_red_reml <- mgcv::gam(red_form(tr), data = dd, method = "REML",
                            weights = w, select = TRUE, na.action = na.exclude)

    # ML versions, needed to compare models with different fixed effects.
    f_full_ml <- mgcv::gam(full_form(tr), data = dd, method = "ML",
                           weights = w, select = TRUE, na.action = na.exclude)
    f_red_ml  <- mgcv::gam(red_form(tr),  data = dd, method = "ML",
                           weights = w, select = TRUE, na.action = na.exclude)

    # Approximate test of the environmental block.
    av <- tryCatch(stats::anova(f_red_ml, f_full_ml, test = "Chisq"),
                   error = function(e) NULL)

    p_env <- NA_real_
    if (!is.null(av)) {
      avd <- tryCatch(as.data.frame(av), error = function(e) NULL)
      if (!is.null(avd)) {
        pcols <- grep("^Pr\\(", colnames(avd), value = TRUE)
        if (length(pcols)) p_env <- as.numeric(tail(avd[[pcols[1]]], 1))
      }
    }

    # Model summaries.
    sm_full <- summary(f_full)
    sm_red  <- summary(f_red_reml)

    r2_full  <- sm_full$r.sq
    r2_red   <- sm_red$r.sq
    dev_full <- sm_full$dev.expl
    st       <- sm_full$s.table

    # Effective degrees of freedom of the environmental smooths.
    env_edf <- if (!is.null(st) && nrow(st)) {
      sum(vapply(env_vars, function(v) {
        keep <- grepl(paste0("^s\\(", v, "(,|\\))"), rownames(st))
        if (any(keep)) sum(st[keep, "edf"]) else 0
      }, numeric(1)))
    } else {
      NA_real_
    }

    # Individual environmental-smooth p-values.
    env_p <- if (!is.null(st) && nrow(st)) {
      out <- numeric(length(env_vars)); names(out) <- env_vars
      for (v in env_vars) {
        keep <- grepl(paste0("^s\\(", v, "(,|\\))"), rownames(st))
        out[v] <- if (any(keep) && "p-value" %in% colnames(st))
          st[keep, "p-value"][1] else NA_real_
      }
      out
    } else {
      setNames(rep(NA_real_, length(env_vars)), env_vars)
    }

    result_row <- data.frame(
      trait = tr,
      n = nrow(dd),
      p_env = p_env,
      delta_R2 = r2_full - r2_red,   # environmental gain (conservative)
      R2_full = r2_full,
      R2_reduced = r2_red,
      dev_expl = dev_full,
      env_edf = env_edf,
      stringsAsFactors = FALSE
    )

    list(df = result_row, model = f_full, model_reduced = f_red_reml, env_p = env_p)
  })

  tab <- do.call(rbind, lapply(rows, `[[`, "df"))
  tab$p_env_fdr <- stats::p.adjust(tab$p_env, method = "BH")

  list(
    table          = tab,
    models         = stats::setNames(lapply(rows, `[[`, "model"),         traits),
    reduced_models = stats::setNames(lapply(rows, `[[`, "model_reduced"), traits),
    env_p          = stats::setNames(lapply(rows, `[[`, "env_p"),         traits)
  )
}
```

```{r}
#| label: gam-run
#| message: false

cat("\n=== Confounder-aware spatial GAMs ===\n")

gam_out <- fit_trait_models(dat)

print(gam_out$table)

write.csv(gam_out$table,
          file.path(out_dir, "gam_env_association.csv"),
          row.names = FALSE)

for (tr in trait_cols) {
  cat("\n-----", tr, "-----\n")
  print(summary(gam_out$models[[tr]]))
}
```

### 4a. Per-driver environmental smooths (table)

The block-level test above says *whether* the environment matters; this table
says *which* driver, with its effective degrees of freedom, F and (BH-adjusted)
p-value. Non-linear structure is captured by the edf (`≈1` ≈ linear, `>1` ≈
curved).

```{r}
#| label: gam-smooths-extract
#| message: false

extract_env_smooths <- function(gam_out, env_vars) {

  out <- lapply(names(gam_out$models), function(tr) {

    m  <- gam_out$models[[tr]]
    sm <- summary(m)$s.table

    if (is.null(sm) || !nrow(sm)) return(NULL)

    vars_present <- env_vars[vapply(env_vars, function(v)
      any(grepl(paste0("^s\\(", v, "(,|\\))"), rownames(sm))), logical(1))]

    if (!length(vars_present)) return(NULL)

    do.call(rbind, lapply(vars_present, function(v) {
      rn <- grep(paste0("^s\\(", v, "(,|\\))"), rownames(sm), value = TRUE)
      x  <- sm[rn[1], , drop = FALSE]
      data.frame(
        trait = tr,
        variable = v,
        edf = x[1, "edf"],
        ref_df = x[1, "Ref.df"],
        F = x[1, "F"],
        p_value = x[1, "p-value"],
        stringsAsFactors = FALSE
      )
    }))
  })

  out <- do.call(rbind, out)

  out$p_FDR <- stats::p.adjust(out$p_value, method = "BH")
  out$interpretation <- ifelse(out$p_FDR < 0.05, "significant smooth", "ns")

  rownames(out) <- NULL
  out
}

env_smooths <- extract_env_smooths(gam_out, continuous_vars)

print(env_smooths)
write.csv(env_smooths,
          file.path(out_dir, "gam_individual_environmental_smooths.csv"),
          row.names = FALSE)

env_smooths[order(env_smooths$trait, env_smooths$p_FDR), ]
```

### 4b. Per-driver environmental smooths (figure)

Partial-effect curves for every environmental driver × trait combination, with
95% confidence ribbons. The panel grid varies the environmental driver down the
rows and the colour trait across the columns, so a driver's effect on all four
traits can be read off a single row.

```{r}
#| label: gam-smooths-plot-fn
#| message: false

plot_environmental_smooths <- function(models,
                                       env_vars,
                                       traits = names(models),
                                       out_svg,
                                       n = 100) {

  if (!requireNamespace("ggplot2", quietly = TRUE))
    stop("Package 'ggplot2' is required.")

  plot_list <- list()

  for (tr in traits) {

    m <- models[[tr]]
    if (is.null(m)) next

    md <- m$model

    for (v in env_vars) {

      if (!v %in% names(md)) next
      if (!is.numeric(md[[v]])) next

      x  <- md[[v]]
      xr <- range(x, na.rm = TRUE)
      if (!all(is.finite(xr)) || diff(xr) == 0) next

      # Which smooth corresponds to this single-term predictor?
      sm_idx <- which(vapply(m$smooth, function(s)
        length(s$term) == 1 && identical(s$term, v), logical(1)))
      if (!length(sm_idx)) next

      sm_name <- m$smooth[[sm_idx[1]]]$label

      # Start from a real model observation, repeat it over the prediction
      # grid, then vary only the environmental predictor.
      newdata <- md[1, , drop = FALSE]
      newdata <- newdata[rep(1, n), , drop = FALSE]
      rownames(newdata) <- NULL
      newdata[[v]] <- seq(xr[1], xr[2], length.out = n)

      pr <- predict(m, newdata = newdata, type = "terms",
                    terms = sm_name, se.fit = TRUE)

      fit <- as.numeric(pr$fit[, 1])
      se  <- as.numeric(pr$se.fit[, 1])

      plot_list[[length(plot_list) + 1]] <- data.frame(
        trait = tr,
        variable = v,
        x = newdata[[v]],
        fit = fit,
        lower = fit - 1.96 * se,
        upper = fit + 1.96 * se,
        stringsAsFactors = FALSE
      )
    }
  }

  if (!length(plot_list)) {
    stop("No environmental smooths could be extracted. ",
         "Check that env_vars match the variables used in the GAMs.")
  }

  plot_data <- do.call(rbind, plot_list)
  rownames(plot_data) <- NULL

  # One facet per environmental-driver x trait combination.
  plot_data$panel <- interaction(plot_data$variable, plot_data$trait,
                                 sep = " | ", drop = TRUE)

  p <- ggplot2::ggplot(plot_data, ggplot2::aes(x = x, y = fit)) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = lower, ymax = upper), alpha = 0.2) +
    ggplot2::geom_line(linewidth = 0.8) +
    ggplot2::geom_hline(yintercept = 0, linetype = 2, linewidth = 0.4) +
    ggplot2::facet_wrap(~ panel, scales = "free", ncol = length(traits)) +
    ggplot2::labs(x = NULL, y = "Partial effect") +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(
      strip.background = ggplot2::element_rect(fill = "white", colour = "black"),
      strip.text = ggplot2::element_text(face = "bold"),
      panel.grid = ggplot2::element_line(colour = "grey90", linewidth = 0.25),
      legend.position = "bottom"
    )

  if (requireNamespace("svglite", quietly = TRUE)) {
    ggplot2::ggsave(out_svg, plot = p, device = svglite::svglite,
                    width = 10, height = 8, units = "in")
  } else {
    ggplot2::ggsave(out_svg, plot = p, device = "svg",
                    width = 10, height = 8, units = "in")
  }

  invisible(p)
}
```

```{r}
#| label: gam-smooths-plot-run
#| fig-width: 12
#| fig-height: 14
#| fig-cap: "Partial effects of each environmental driver on each colour trait (GAM smooths, 95% CI)."

plot_environmental_smooths(
  models  = gam_out$models,
  env_vars = continuous_vars,
  traits  = trait_cols,
  out_svg = file.path(out_dir, "gam_environmental_smooths.svg")
)
```

### 4c. Combined GAM summary

One tidy table joining the block-level environmental test with the per-driver
smooth results, written for the manuscript.

```{r}
#| label: gam-summary
#| message: false

gam_summary <- merge(
  gam_out$table[, c("trait", "n", "p_env", "p_env_fdr",
                    "delta_R2", "R2_full", "dev_expl")],
  env_smooths[, c("trait", "variable", "edf", "F", "p_value", "p_FDR")],
  by = "trait"
)

write.csv(gam_summary,
          file.path(out_dir, "gam_environmental_results.csv"),
          row.names = FALSE)

gam_summary
```

### 4d. Crystallize GAM: EnZ, elevation, subspecies

Synthesis of the §4 exploration: two crystallised traits (`contrast`,
`warmth`) modelled as a function of Environmental Zone (`EnZ_name`) and
elevation, controlling for subspecies, season (`s(doy, bs = "cc")`) and space
(`s(x, y)` on EPSG:3035 coordinates). Like §4, each trait gets a **full** and
a **reduced** model (reduced drops `EnZ_name + s(elevation)`); the
environmental block is tested with an approximate Chi-square LRT on **ML**
fits, while R² / deviance explained come from the **REML** fits. Quality
weights `w` and `select = TRUE` match §4. Nothing here overwrites `dat` —
all 4d modelling uses a copy `dat_enz` — so §5 still sees the original data.

```{r}
#| label: gam-enz-maps
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Environmental Zones (EnZ) and Strata (EnS) of Europe with observation locations."
#| fig-subcap:
#|   - "Environmental Zones of Europe."
#|   - "Environmental Zones and observation locations."
#|   - "Environmental Strata and observation locations."
#| layout-ncol: 1

library(emmeans)
library(gratia)

ens <- sf::st_read(file.path(
  ENVROOT,
  "Metzger_Environmental_stratification_of_Europe(EnZ)/ens_v8.shp"
))

enz_map <- ens %>%
  group_by(EnZ, EnZ_name) %>%
  summarise(geometry = st_union(geometry), .groups = "drop")

ggplot(enz_map) +
  geom_sf(aes(fill = EnZ_name), colour = "grey30", linewidth = 0.2) +
  theme_minimal() +
  labs(
    title = "Environmental Zones of Europe",
    fill = "EnZ"
  )

ggplot() +
  geom_sf(
    data = enz_map,
    aes(fill = EnZ_name),
    colour = "grey40",
    linewidth = 0.2
  ) +
  geom_sf(
    data = occ_sf_full,
    colour = "black",
    size = 1,
    alpha = 0.6
  ) +
  theme_minimal() +
  labs(
    title = "Environmental Zones and Observation Locations",
    fill = "EnZ"
  )

ggplot() +
  geom_sf(
    data = ens,
    aes(fill = EnS_name),
    colour = "grey50",
    linewidth = 0.1
  ) +
  geom_sf(
    data = occ_sf_full,
    colour = "black",
    size = 1,
    alpha = 0.6
  ) +
  theme_minimal() +
  labs(
    title = "Environmental Strata and Observation Locations",
    fill = "EnS"
  )
```

```{r}
#| label: gam-enz-fit
#| message: false

# Copy so the global `dat` (used in §5) is never modified.
dat_enz <- dat %>%
  mutate(
    EnZ_name = factor(EnZ_name)
  )

pts_enz <- sf::st_as_sf(
  dat_enz,
  coords = c("longitude", "latitude"),
  crs = 4326
)

pts_enz <- sf::st_transform(pts_enz, 3035)

xy_enz <- sf::st_coordinates(pts_enz)

dat_enz$x <- xy_enz[, 1]
dat_enz$y <- xy_enz[, 2]

table(dat_enz$EnZ_name)

m_contrast_EnZ <- mgcv::gam(
  contrast ~
    EnZ_name +
    subspecies +
    s(elevation) +
    s(doy, bs = "cc", k = 10) +
    s(x, y, k = 30),
  data = dat_enz,
  method = "REML",
  weights = w,
  select = TRUE,
  knots = list(doy = c(0.5, 365.5))
)

m_warmth_EnZ <- mgcv::gam(
  warmth ~
    EnZ_name +
    subspecies +
    s(elevation) +
    s(doy, bs = "cc", k = 10) +
    s(x, y, k = 30),
  data = dat_enz,
  method = "REML",
  weights = w,
  select = TRUE,
  knots = list(doy = c(0.5, 365.5))
)

# Reduced models: everything except the EnZ + elevation block.
m_contrast_EnZ_red <- mgcv::gam(
  contrast ~
    subspecies +
    s(doy, bs = "cc", k = 10) +
    s(x, y, k = 30),
  data = dat_enz,
  method = "REML",
  weights = w,
  select = TRUE,
  knots = list(doy = c(0.5, 365.5))
)

m_warmth_EnZ_red <- mgcv::gam(
  warmth ~
    subspecies +
    s(doy, bs = "cc", k = 10) +
    s(x, y, k = 30),
  data = dat_enz,
  method = "REML",
  weights = w,
  select = TRUE,
  knots = list(doy = c(0.5, 365.5))
)

# ML refits for the block test (REML is invalid with different fixed effects).
enz_block_test <- function(full_reml, red_reml, trait, full_form, red_form, d) {
  full_ml <- mgcv::gam(full_form, data = d, method = "ML",
                       weights = w, select = TRUE,
                       knots = list(doy = c(0.5, 365.5)))
  red_ml  <- mgcv::gam(red_form,  data = d, method = "ML",
                       weights = w, select = TRUE,
                       knots = list(doy = c(0.5, 365.5)))
  av <- tryCatch(stats::anova(red_ml, full_ml, test = "Chisq"),
                 error = function(e) NULL)
  p_env <- NA_real_
  if (!is.null(av)) {
    avd <- tryCatch(as.data.frame(av), error = function(e) NULL)
    if (!is.null(avd)) {
      pcols <- grep("^Pr\\(", colnames(avd), value = TRUE)
      if (length(pcols)) p_env <- as.numeric(tail(avd[[pcols[1]]], 1))
    }
  }
  sm_full <- summary(full_reml)
  sm_red  <- summary(red_reml)
  data.frame(
    trait = trait,
    n = nrow(d),
    p_env = p_env,
    delta_R2 = sm_full$r.sq - sm_red$r.sq,
    R2_full = sm_full$r.sq,
    R2_reduced = sm_red$r.sq,
    dev_expl = sm_full$dev.expl,
    stringsAsFactors = FALSE
  )
}

contrast_full_form <- contrast ~ EnZ_name + subspecies + s(elevation) +
  s(doy, bs = "cc", k = 10) + s(x, y, k = 30)
contrast_red_form  <- contrast ~ subspecies +
  s(doy, bs = "cc", k = 10) + s(x, y, k = 30)
warmth_full_form   <- warmth ~ EnZ_name + subspecies + s(elevation) +
  s(doy, bs = "cc", k = 10) + s(x, y, k = 30)
warmth_red_form    <- warmth ~ subspecies +
  s(doy, bs = "cc", k = 10) + s(x, y, k = 30)

gam_enz_block <- rbind(
  enz_block_test(m_contrast_EnZ, m_contrast_EnZ_red, "contrast",
                 contrast_full_form, contrast_red_form, dat_enz),
  enz_block_test(m_warmth_EnZ, m_warmth_EnZ_red, "warmth",
                 warmth_full_form, warmth_red_form, dat_enz)
)
gam_enz_block$p_env_fdr <- stats::p.adjust(gam_enz_block$p_env, method = "BH")

print(gam_enz_block)
write.csv(gam_enz_block,
          file.path(out_dir, "gam_EnZ_block_test.csv"),
          row.names = FALSE)
```

Per-driver EnZ effects (table): estimated marginal means per zone with Tukey
pairwise contrasts, plus the elevation-smooth edf / F / p-value — the analogue
of §4a.

```{r}
#| label: gam-enz-table
#| message: false

# --- Contrast ---
summary(m_contrast_EnZ)
mgcv::gam.check(m_contrast_EnZ)

enz_contrast <- emmeans::emmeans(
  m_contrast_EnZ,
  ~ EnZ_name
)
enz_contrast
pairs(enz_contrast, adjust = "tukey")

# --- Warmth ---
summary(m_warmth_EnZ)
mgcv::gam.check(m_warmth_EnZ)

enz_warmth <- emmeans::emmeans(
  m_warmth_EnZ,
  ~ EnZ_name
)
enz_warmth
pairs(enz_warmth, adjust = "tukey")

# Elevation smooths, mirroring the §4a per-driver table.
enz_elev_smooths <- do.call(rbind, lapply(
  list(contrast = m_contrast_EnZ, warmth = m_warmth_EnZ),
  function(m) {
    st <- summary(m)$s.table
    rn <- grep("^s\\(elevation(,|\\))", rownames(st), value = TRUE)
    data.frame(
      edf = st[rn[1], "edf"],
      ref_df = st[rn[1], "Ref.df"],
      F = st[rn[1], "F"],
      p_value = st[rn[1], "p-value"]
    )
  }
))
enz_elev_smooths$trait <- rownames(enz_elev_smooths)
rownames(enz_elev_smooths) <- NULL
enz_elev_smooths$variable <- "elevation"
enz_elev_smooths$p_FDR <- stats::p.adjust(enz_elev_smooths$p_value, method = "BH")
print(enz_elev_smooths)

# Tidy emmeans for the figure and the manuscript table.
enz_contrast_df <- as.data.frame(enz_contrast)
enz_warmth_df   <- as.data.frame(enz_warmth)
enz_emmeans_tidy <- rbind(
  data.frame(trait = "contrast", enz_contrast_df, stringsAsFactors = FALSE),
  data.frame(trait = "warmth",   enz_warmth_df,   stringsAsFactors = FALSE)
)
print(enz_emmeans_tidy)
write.csv(enz_emmeans_tidy,
          file.path(out_dir, "gam_EnZ_emmeans.csv"),
          row.names = FALSE)
write.csv(enz_elev_smooths,
          file.path(out_dir, "gam_EnZ_elevation_smooths.csv"),
          row.names = FALSE)
```

Per-driver EnZ effects (figure): marginal-mean point–ranges per zone with 95%
confidence limits — the analogue of §4b.

```{r}
#| label: gam-enz-figure
#| message: false
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Colour contrast and warmth across Environmental Zones (estimated marginal means, 95% CI)."
#| fig-subcap:
#|   - "Colour contrast across environmental zones."
#|   - "Colour warmth across environmental zones."
#| layout-ncol: 2

ggplot(enz_contrast_df,
       aes(x = EnZ_name, y = emmean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.15
  ) +
  labs(
    x = "Environmental zone",
    y = "Estimated contrast",
    title = "Colour contrast across environmental zones"
  ) +
  theme_classic()

ggplot(enz_warmth_df,
       aes(x = EnZ_name, y = emmean)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.15
  ) +
  labs(
    x = "Environmental zone",
    y = "Estimated warmth",
    title = "Colour warmth across environmental zones"
  ) +
  theme_classic()
```

Elevation and spatial partial effects for the crystallised models.

```{r}
#| label: gam-enz-figure-smooths
#| message: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Elevation (first smooth) and spatial s(x, y) partial effects for the crystallised EnZ models."
#| fig-subcap:
#|   - "Contrast: elevation smooth."
#|   - "Contrast: spatial smooth."
#|   - "Warmth: elevation smooth."
#|   - "Warmth: spatial smooth."
#| layout-ncol: 2

plot(
  m_contrast_EnZ,
  select = 1,
  shade = TRUE,
  residuals = FALSE
)

gratia::draw(
  m_contrast_EnZ,
  select = "s(x,y)"
)

plot(
  m_warmth_EnZ,
  select = 1,
  shade = TRUE,
  residuals = FALSE
)

gratia::draw(
  m_warmth_EnZ,
  select = "s(x,y)"
)
```

Combined 4d summary for the manuscript — the analogue of §4c.

```{r}
#| label: gam-enz-summary
#| message: false

gam_EnZ_summary <- merge(
  gam_enz_block[, c("trait", "n", "p_env", "p_env_fdr",
                    "delta_R2", "R2_full", "dev_expl")],
  enz_elev_smooths[, c("trait", "variable", "edf", "F", "p_value", "p_FDR")],
  by = "trait"
)

write.csv(gam_EnZ_summary,
          file.path(out_dir, "gam_EnZ_crystallized_results.csv"),
          row.names = FALSE)

gam_EnZ_summary
```

Are light / high-contrast / warm snakes non-randomly distributed in space?

A Mantel test only picks up *gradual* distance-decay gradients; a "checkerboard"
of localised patches (light snakes in one valley but not the adjacent one) is
invisible to matrix correlations. We therefore test each trait against a spatial
null model by **permuting trait values over the observation locations**
(`PERM` times) and recomputing, on a k-nearest-neighbour graph between
observations:

* **Moran's *I*** — clustering of similar values (`I > 0`) or alternation of
  contrasting values (`I < 0`) at the local scale;
* **neighbour contrast** — mean `|trait_i − trait_j|` among neighbours, elevated
  under checkerboard-like localised variation regardless of any gradient.

Permutation p-values compare the observed statistic's distance from the
*estimated null centre* (not an assumed zero), and BH correction is applied
across all trait × statistic tests together.

```{r}
#| label: spatial-test-fns
#| message: false

# k-nearest-neighbour graph on a local equirectangular approximation.
spatial_nb <- function(d, k = 50) {
  n <- nrow(d)
  k <- min(k, n - 1)
  lon <- d$longitude; lat <- d$latitude
  latm <- mean(lat) * pi / 180
  D <- sqrt((outer(lon, lon, "-") * 111.32 * cos(latm))^2 +
              (outer(lat, lat, "-") * 110.574)^2)
  diag(D) <- Inf
  nn <- apply(D, 1, order)[seq_len(k), , drop = FALSE]   # k x n: column = site
  list(i = rep(seq_len(n), each = k),
       j = as.vector(nn),
       w = rep(1 / k, n * k))
}

spatial_randomness_test <- function(d, traits = trait_cols, k = 6,
                                    perms = PERM, seed = 42) {

  if (nrow(d) < 10) stop("spatial test: need >= 10 observations.")

  gr <- spatial_nb(d, k = k)

  out <- lapply(traits, function(tr) {

    z <- as.numeric(d[[tr]])
    if (anyNA(z)) stop(sprintf("Trait '%s' contains missing values.", tr))

    zbar  <- mean(z)
    zc    <- z - zbar
    denom <- sum(zc^2)

    # --- Observed statistics -------------------------------------------------
    obs_I <- if (denom > 0) sum(gr$w * zc[gr$i] * zc[gr$j]) / denom else 0
    obs_D <- mean(abs(z[gr$i] - z[gr$j]))

    # --- Permutation distributions ------------------------------------------
    set.seed(seed + which(traits == tr))
    perm_I <- numeric(perms)
    perm_D <- numeric(perms)

    for (r in seq_len(perms)) {
      zs  <- sample(z)          # redistribute trait values over sites
      zcs <- zs - zbar

      perm_I[r] <- if (denom > 0) sum(gr$w * zcs[gr$i] * zcs[gr$j]) / denom else 0
      perm_D[r] <- mean(abs(zs[gr$i] - zs[gr$j]))
    }

    # --- Null centres estimated from the permutations -----------------------
    # The k-NN graph is directed/asymmetric and the statistic is a simplified
    # Moran's I, so we do not assume the null is centred at zero.
    centre_I <- mean(perm_I)
    centre_D <- mean(perm_D)

    # --- One-sided p-values --------------------------------------------------
    pI_upper <- (sum(perm_I >= obs_I) + 1) / (perms + 1)
    pI_lower <- (sum(perm_I <= obs_I) + 1) / (perms + 1)
    pD_upper <- (sum(perm_D >= obs_D) + 1) / (perms + 1)
    pD_lower <- (sum(perm_D <= obs_D) + 1) / (perms + 1)

    # --- Coherent two-sided p-values ----------------------------------------
    pI_two <- (sum(abs(perm_I - centre_I) >= abs(obs_I - centre_I)) + 1) / (perms + 1)
    pD_two <- (sum(abs(perm_D - centre_D) >= abs(obs_D - centre_D)) + 1) / (perms + 1)

    # --- Direction of the observed effect -----------------------------------
    direction_I <- if (obs_I > centre_I) "positive" else if (obs_I < centre_I) "negative" else "none"
    direction_D <- if (obs_D > centre_D) "high"     else if (obs_D < centre_D) "low"      else "none"

    data.frame(
      trait = tr,
      statistic = c("Moran_I", "neighbour_diff"),
      observed = c(obs_I, obs_D),
      null_mean = c(centre_I, centre_D),
      p_perm = c(pI_two, pD_two),
      p_lower = c(pI_lower, pD_lower),
      p_upper = c(pI_upper, pD_upper),
      direction = c(direction_I, direction_D),
      stringsAsFactors = FALSE
    )
  })

  res <- do.call(rbind, out)
  res$p_FDR <- stats::p.adjust(res$p_perm, method = "BH")

  # --- Plain-language interpretation ----------------------------------------
  res$interpretation <- "ns"

  idx_I <- res$statistic == "Moran_I" & res$p_FDR < 0.05
  res$interpretation[idx_I & res$direction == "positive"] <-
    "positive spatial autocorrelation (similar neighbours)"
  res$interpretation[idx_I & res$direction == "negative"] <-
    "negative spatial autocorrelation (contrasting neighbours)"

  idx_D <- res$statistic == "neighbour_diff" & res$p_FDR < 0.05
  res$interpretation[idx_D & res$direction == "low"] <-
    "local similarity (neighbours more similar than expected)"
  res$interpretation[idx_D & res$direction == "high"] <-
    "local contrast (neighbours more different than expected)"

  rownames(res) <- NULL
  res
}
```

```{r}
#| label: spatial-test-run
#| message: false

cat("\n=== Non-random spatial arrangement of colour ===\n")
cat(sprintf("  (k-NN graph k=%d, %d permutations per trait; two-sided p, BH-FDR)\n",
            min(6, nrow(dat) - 1), PERM))

sp_test <- spatial_randomness_test(dat, k = 20, perms = 9999)
print(sp_test)

write.csv(sp_test, file.path(out_dir, "spatial_randomness_tests.csv"),
          row.names = FALSE)
```

### 5a. Spatial arrangement maps

Local Moran-scatterplot classification. Each point is coloured by the sign
combination of its own standardised trait value (*x*) and the mean of its
k nearest neighbours (*lag*):

* **HH** high–high and **LL** low–low → spatial clustering;
* **HL** high–low and **LH** low–high → local spatial outliers.

Point size grows with distance from the trait mean. The panel annotation
reports the global Moran's *I*, its FDR and the direction from the permutation
test.

```{r}
#| label: spatial-maps-fn
#| message: false

plot_spatial_arrangement_maps <- function(
    d,
    traits = trait_cols,
    k = 6,
    sp_test = NULL,
    out_svg,
    aggregate_km = 0
) {

  if (!all(c("longitude", "latitude") %in% names(d)))
    stop("longitude and latitude columns are required.")

  if (nrow(d) < 2)
    stop("Need at least 2 observations.")

  if (aggregate_km < 0)
    stop("aggregate_km must be >= 0.")

  n <- nrow(d)
  k_use <- min(k, n - 1)

  gr <- spatial_nb(d, k = k_use)

  cl_cols <- c(
    HH = "#d73027",
    LL = "#4575b4",
    HL = "#f1eeea",
    LH = "#eff1f2"
  )

  # --------------------------------------------------------------
  # Calculate Moran scatterplot classes for each trait
  # --------------------------------------------------------------

  plot_list <- lapply(traits, function(tr) {

    z_raw <- as.numeric(d[[tr]])

    if (anyNA(z_raw))
      stop(sprintf("Trait '%s' contains missing values.", tr))

    z <- as.numeric(scale(z_raw))

    lag <- numeric(n)
    for (ii in seq_len(n)) {
      lag[ii] <- mean(z[gr$j[gr$i == ii]])
    }

    cl <- rep(NA_character_, n)
    cl[z > 0 & lag > 0] <- "HH"
    cl[z < 0 & lag < 0] <- "LL"
    cl[z > 0 & lag < 0] <- "HL"
    cl[z < 0 & lag > 0] <- "LH"
    cl[z == 0 & lag > 0] <- "LH"
    cl[z == 0 & lag < 0] <- "LL"
    cl[z > 0 & lag == 0] <- "HH"
    cl[z < 0 & lag == 0] <- "HL"

    data.frame(
      trait = tr,
      longitude = d$longitude,
      latitude = d$latitude,
      z = z,
      lag = lag,
      cluster = factor(cl, levels = names(cl_cols)),
      stringsAsFactors = FALSE
    )
  })

  plot_data <- do.call(rbind, plot_list)

  # --------------------------------------------------------------
  # Aggregate nearby sites for visualization only
  # --------------------------------------------------------------

  aggregate_points <- function(x, distance_km) {

    if (distance_km == 0) {
      return(data.frame(
        longitude = x$longitude,
        latitude  = x$latitude,
        cluster   = x$cluster,
        n_sites   = 1L,
        mean_absz = abs(x$z)
      ))
    }

    out <- list()
    counter <- 0

    for (cl in levels(x$cluster)) {

      sub <- x[!is.na(x$cluster) & x$cluster == cl, , drop = FALSE]

      if (!nrow(sub)) next

      if (nrow(sub) == 1) {
        counter <- counter + 1
        out[[counter]] <- data.frame(
          longitude = sub$longitude,
          latitude  = sub$latitude,
          cluster   = cl,
          n_sites   = 1L,
          mean_absz = abs(sub$z)
        )
        next
      }

      latm <- mean(sub$latitude) * pi / 180
      D <- sqrt(
        (outer(sub$longitude, sub$longitude, "-") * 111.32 * cos(latm))^2 +
        (outer(sub$latitude,  sub$latitude,  "-") * 110.574)^2
      )
      diag(D) <- 0

      hc <- stats::hclust(stats::as.dist(D), method = "complete")
      groups <- stats::cutree(hc, h = distance_km)

      for (g in sort(unique(groups))) {
        ii <- which(groups == g)
        counter <- counter + 1
        out[[counter]] <- data.frame(
          longitude = mean(sub$longitude[ii], na.rm = TRUE),
          latitude  = mean(sub$latitude[ii],  na.rm = TRUE),
          cluster   = cl,
          n_sites   = length(ii),
          mean_absz = mean(abs(sub$z[ii]), na.rm = TRUE)
        )
      }
    }

    do.call(rbind, out)
  }

  aggregated_list <- lapply(
    split(plot_data, plot_data$trait),
    aggregate_points,
    distance_km = aggregate_km
  )

  aggregated <- do.call(
    rbind,
    lapply(names(aggregated_list), function(tr) {
      x <- aggregated_list[[tr]]
      x$trait <- tr
      x
    })
  )

  rownames(aggregated) <- NULL

  # --------------------------------------------------------------
  # Annotation
  # --------------------------------------------------------------

  annotations <- data.frame(
    trait = traits,
    label = paste0("k-NN = ", k_use),
    stringsAsFactors = FALSE
  )

  if (!is.null(sp_test)) {
    for (tr in traits) {
      sub <- sp_test[
        sp_test$trait == tr & sp_test$statistic == "Moran_I",
        , drop = FALSE
      ]
      if (nrow(sub)) {
        ptxt <- if (sub$p_FDR[1] <= 0.001) "FDR < 0.001" else sprintf("FDR = %.3f", sub$p_FDR[1])
        direction_txt <- if (sub$direction[1] == "positive") {
          "Positive spatial autocorrelation"
        } else if (sub$direction[1] == "negative") {
          "Negative spatial autocorrelation"
        } else {
          "No directional spatial autocorrelation"
        }
        annotations$label[annotations$trait == tr] <- paste0(
          "Moran's I = ", sprintf("%.2f", sub$observed[1]), "; ", ptxt, "\n", direction_txt
        )
      }
    }
  }

  # --------------------------------------------------------------
  # Map extent
  # --------------------------------------------------------------

  x_range <- range(d$longitude, na.rm = TRUE)
  y_range <- range(d$latitude,  na.rm = TRUE)
  x_pad <- max(diff(x_range) * 0.05, 0.1)
  y_pad <- max(diff(y_range) * 0.05, 0.1)
  x_limits <- c(x_range[1] - x_pad, x_range[2] + x_pad)
  y_limits <- c(y_range[1] - y_pad, y_range[2] + y_pad)

  # --------------------------------------------------------------
  # World map
  # --------------------------------------------------------------

  world <- NULL
  if (requireNamespace("maps", quietly = TRUE)) {
    world <- ggplot2::map_data("world")
    world <- world[
      world$long >= x_limits[1] & world$long <= x_limits[2] &
      world$lat  >= y_limits[1] & world$lat  <= y_limits[2],
      , drop = FALSE
    ]
  }

  # --------------------------------------------------------------
  # Plot
  # --------------------------------------------------------------

  p <- ggplot2::ggplot(aggregated, ggplot2::aes(x = longitude, y = latitude))

  if (!is.null(world) && nrow(world) > 0) {
    p <- p +
      ggplot2::geom_polygon(
        data = world,
        ggplot2::aes(x = long, y = lat, group = group),
        inherit.aes = FALSE,
        fill = "grey95", colour = "grey75", linewidth = 0.25
      )
  }

  p <- p +
    ggplot2::geom_point(
      ggplot2::aes(fill = cluster, size = mean_absz, stroke = n_sites),
      shape = 21, colour = "grey20") +
    ggplot2::scale_size_continuous(aesthetics = "stroke", range = c(0.2, 1), name = "Sites merged") +
    ggplot2::scale_fill_manual(
      values = cl_cols,
      drop = FALSE,
      labels = c(HH = "HH: high\u2013high", LL = "LL: low\u2013low",
                 HL = "HL: high\u2013low", LH = "LH: low\u2013high"),
      name = "Moran scatterplot class"
    ) +
    ggplot2::scale_size_continuous(
      range = c(1.5, 6),
      name = "Mean |z| (distance from trait mean)"
    ) +
    ggplot2::facet_wrap(~ trait, ncol = 2) +
    ggplot2::coord_cartesian(xlim = x_limits, ylim = y_limits, expand = FALSE) +
    ggplot2::labs(x = "Longitude", y = "Latitude") +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(
      strip.background = ggplot2::element_rect(fill = "white", colour = "black"),
      strip.text = ggplot2::element_text(face = "bold"),
      panel.grid = ggplot2::element_line(colour = "grey90", linewidth = 0.25),
      legend.position = "bottom",
      legend.box = "vertical",
      plot.margin = ggplot2::margin(8, 8, 8, 8)
    ) +
    ggplot2::geom_text(
      data = unique(merge(annotations, data.frame(trait = traits))),
      ggplot2::aes(x = -Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = -0.02, vjust = 1.15, size = 3.0, lineheight = 1.05
    )

  # --------------------------------------------------------------
  # Save SVG
  # --------------------------------------------------------------

  if (requireNamespace("svglite", quietly = TRUE)) {
    ggplot2::ggsave(filename = out_svg, plot = p, device = svglite::svglite,
                     width = 9, height = 8, units = "in")
  } else {
    ggplot2::ggsave(filename = out_svg, plot = p, device = "svg",
                     width = 9, height = 8, units = "in")
  }

  invisible(p)
}

```

```{r}
plot_spatial_arrangement_maps(
  dat,
  k = 10,
  aggregate_km = 0,
  sp_test = sp_test,
  out_svg = file.path(
    out_dir,
    "spatial_arrangement_k10.svg"
  )
)
```
------------------------------------------------------------------------

## 6. Outputs and headline numbers

All tables and figures are written to `MLM_results/`. The final cells print the
headline numbers used in the text.

```{r}
#| label: outputs-summary

cat(sprintf("\nResults written to: %s\n", out_dir))
cat("\n--- Headline numbers ---\n")

cat("GAM per-trait environmental block (p, FDR, delta R2):\n")
print(gam_out$table[, c("trait", "p_env", "p_env_fdr", "delta_R2")])

cat("\nSpatial randomness per trait (Moran's I & neighbour contrast; p, FDR):\n")
print(sp_test[, c("trait", "statistic", "observed", "p_FDR", "interpretation")],
      row.names = FALSE)
```

```{r}
#| label: files-written

list.files(out_dir)
```

```{r}
#| label: session-info
#| echo: false

sessionInfo()
```
